From correlation to causality: statistical approaches to learning regulatory relationships in large-scale biomolecular investigations

### 2015-09-27 write_up.tex
added last set of edits prior to submission

Figure \ref{mapkInference}A demonstrates the causal inference workflow starting with analysis of statistical associations in the data.  In step 1, a correlation graph between cascade components Raf, Mek, and Erk is assembled from the measurements of protein concentration.  Step 2 reduces the correlation graph to a sparse graph of conditional dependencies (Raf--Mek, and Mek--Erk).  Step 3 interrogates this graph to find putative causal relationships (Raf-->Mek, and Mek-->Erk).  While step 1 has little requirements, step 2 requires replicate biological samples, and step 3 requires systematic interventions (e.g. with protein inhibitors).

@@

The direct causal events can be distinguished from other undesirable types of associations by controlling for (i.e., by holding constant) the intermediaries and the common regulators. The absence of statistical association between two proteins, when the intermediates or the common regulators are controlled at a fixed level, is called {\it conditional independence}.  Causal inference searches for empirical evidence of conditional independence, to distinguish causal events from indirect associations.

Let's see how this applies to the MAPK signaling cascade. We can examine the nature of statistical association between Raf and Erk.  Figure \ref{mapkInference}C compares Raf to Erk, indicating in circles the biological samples where concentrations of Mek are high (here, set to the top quartile).  Note that when we subset the data to only the samples with high Mek (in statistical language, when we condition on Mek being high), we can no longer detect the association between Raf and Erk. This pattern has a mechanistic explanation. As can be seen in Figure \ref{mapk}, the abundance of Erk in each sample is determined by Mek. Therefore, when the abundance of Mek is fixed, Raf does not exert any additional influence on the variability of Erk.  This phenomenon -- the disappearence of the association between Raf and Erk upon conditioning on Mek - is evidence that  Raf and Erk are conditionally independent.  Once conditional independence is inferred from the data, the correlation edge arising from Raf--Erk dependence is removed, resulting in the middle graph of Figure \ref{mapk}A.

@@

\item \textit{Use prior knowledge.} The prior knowledge improves the search for conditional independence. The prior knowledge can be in form of known canonical networks, extracted, e.g. from pathway databases such as KEGG. One example of such prior information is the MAPK pathway. The prior information reduces the search space of unknown associations that need to be considered, enables a more effective use of the data, and increases the confidence in newly discovered statistical associations.  Another example of prior knowledge is contextual information, such as spatial or temporal annotations of the quantitative measurements in the cell. The contextual information can be extracted from the literature or from other complementary (and potentially noisy) datasets. The causal inference algorithms can be extended to weigh evidence of conditional dependence, depending on whether the analytes share a same or temporal context.

>>>>

Figure \ref{mapkInference}A demonstrates the causal inference workflow starting with analysis of statistical associations in the data.  In step 1, a correlation graph between cascade components Raf, Mek, and Erk is assembled from the measurements of protein concentration.  Step 2 reduces the correlation graph to a sparse graph of conditional dependencies (Raf--Mek, and Mek--Erk).  Step 3 interrogates this graph to find putative causal relationships (Raf-->Mek, and Mek-->Erk).  While step 1 has little requirements, step 2 requires multiple samples or replicates, and step 3 requires systematic interventions (e.g. with protein inhibitors).

@@

The direct causal events can be distinguished from other undesirable types of associations by controlling for the intermediaries and the common regulators (e.g, by selecting measurements where they have certain values). The absence of statistical association between two proteins, when the intermediates or the common regulators are controlled at a fixed level, is called {\it conditional independence}.  Causal inference searches for empirical evidence of conditional independence, to distinguish causal events from indirect associations.

Let's see how this applies to the MAPK signaling cascade. We can examine the nature of statistical association between Raf and Erk.  Figure \ref{mapkInference}C compares Raf to Erk, indicating in circles the biological samples where concentrations of Mek are high (here, set to the top quartile).  {\it Note that when we subset the data to only the samples with high Mek (in statistical language, when we condition on Mek being high), we can no longer detect the association between Raf and Erk}. This pattern has a mechanistic explanation. As can be seen in Figure \ref{mapk}, the abundance of Erk in each sample is determined by Mek. Therefore, when the abundance of Mek is fixed, Raf does not exert any additional influence on the variability of Erk.  This phenomenon -- the disappearence of the association between Raf and Erk upon conditioning on Mek - is evidence that  Raf and Erk are conditionally independent.  Once conditional independence is inferred from the data, the correlation edge arising from Raf--Erk dependence is removed, resulting in the middle graph of Figure \ref{mapk}A.

@@

\item \textit{Use prior knowledge.} Prior knowledge improves the search for conditional independence and helps to determine causality. The prior knowledge can be in form of known canonical networks, extracted, e.g. from pathway databases such as KEGG. One example of such prior information is the MAPK pathway. The prior information reduces the search space of unknown associations that need to be considered, enables a more effective use of the data, and increases the confidence in newly discovered statistical associations.  Another example of prior knowledge is contextual information, such as spatial or temporal annotations of the quantitative measurements in the cell. The contextual information can be extracted from the literature or from other complementary (and potentially noisy) datasets. The causal inference algorithms can be extended to weigh evidence of conditional dependence, depending on whether the analytes share a same or temporal context.
